---
title: "ch13_non-centred_practice"
output: html_notebook
---
13M3. Re-estimate the basic Reed frog varying intercept model, but now using a 
Cauchy distribution in place of the Gaussian distribution for the varying 
intercepts. That is, fit this model:

$s_i \sim Binomial(n_i , p_i) \\
logit(p_i) = \alpha_{TANK[i]} \\
\alpha_{[TANK]} \sim Cauchy(\alpha, \sigma) \\
\alpha \sim \mathcal{N}(0, 1) \\
\alpha \sim Exponential(1)$

(You are likely to see many divergent transitions for this model. Can you 
figure out why? Can you fix them?) Compare the posterior means of the 
intercepts, α tank , to the posterior means produced in the chapter, using the 
customary Gaussian prior. Can you explain the pattern of differences? Take 
note of any change in the mean α as well.
```{r results="hide"}
library(rethinking)
data("reedfrogs")
d <- reedfrogs
str(d)

d$tank <- 1:nrow(d)

dat <- list(
  S = as.integer(d$surv),
  pred = as.integer(d$pred),
  big=as.integer(d$size),
  N=as.integer(d$density),
  tank=as.integer(d$tank)
)

m13m3 <- ulam(
  alist(
    S ~ dbinom(N, p),
    logit(p) <- a[tank],
    a[tank] ~ dcauchy(alpha, sigma),
    alpha ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ), data=dat, chains=4, log_lik = TRUE
)
```


from the chapter

```{r results="hide"}
m13.2 <- ulam(
  alist(
    S ~ dbinom( N , p ),
    logit(p) <- a[tank],
    a[tank] ~ dnorm(a_bar, sigma),
    a_bar ~ dnorm(0 , 1.5),
    sigma ~ dexp(1)
  ), data=dat, chains=4, log_lik = TRUE
)
```
```{r}
precis_plot(precis(m13.2, depth=2), xlim = c(-2,20))
precis_plot(precis(m13m3, depth=2), xlim=c(-2,20))
precis(m13.2)
precis(m13m3)
```


basically those a_tanks that were already quite high are now allowed to grow
crazily big. Unsure as to what the cauchy distirbution is. It's learnt a
slightly higher alpha bar now, and also oddly enough a lower sigma, despite
this being far bigger variation in those crazy alphas

to be honest though, the large individual alphas are all basically "always"
on the log-odds scale already before cauchy.

Apparently Cauchy has an undefined mean cos of tose crazy tails! That's where
we get those large a values from
```{r}
m13m3nc <- ulam(
  alist(
    S ~ dbinom(N, p),
    logit(p) <- a_bar + z[tank]*sigma_a,
    z[tank] ~ dcauchy(0, 1),
    a_bar ~ dnorm(0, 1),
    sigma_a ~ dexp(1),
    gq > vector [tank]:a <<- a_bar + z*sigma_a
  ), data=dat, chains=4, log_lik = TRUE
)
compare(m13m3m, m13m3nc)
precis(m13m3, m13m3nc)
```
re-parameterised to have no divergent transitions!